---
title: "Train exploration"
author: "JBO"
date: "21 January 2017"
output: html_document
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = TRUE)
library(jsonlite)
library(tidyverse)
library(lubridate)
library(viridis)

# to fix issues with directories
library(rprojroot)
root <- rprojroot::is_rstudio_project

```


##Load data
```{r, warning=FALSE, message=FALSE}
con <- url("http://gtfs.irail.be/nmbs/feedback/occupancy-until-20161029.newlinedelimitedjsonobjects")
train_data <- as_data_frame(jsonlite::flatten(jsonlite::stream_in(con), recursive=TRUE))%>%
  setNames(gsub(x = names(.), pattern = 'post.', replacement = ''))%>%
  select(-user_agent, -querytype, -date)%>%
  mutate(querytime = ymd_hms(querytime, tz = 'CET'))%>%
  mutate(occupancy = gsub(occupancy, pattern = 'http://api.irail.be/terms/', replacement = ''))

station_data <- read_csv(paste0(root$find_file("data/"),"/stations.csv"))%>%
  setNames(tolower(make.names(names(.))))%>%
  # remove french translation after / from name
  mutate(name = gsub(name, pattern = '/.*$', replacement = ''))
```


##How often is each station mentioned in the data?
```{r, warning=FALSE}
station_frequency <- train_data%>%
  gather(key = direction, value = station, c(from, to))%>%
  inner_join(station_data%>%
               select(uri, name),
             by = c('station' = 'uri'))%>%
  group_by(name, direction)%>%
  summarise(n_records = n())%>%
  mutate(total_mentions = sum(n_records))%>%
  ungroup()%>%
  arrange(desc(total_mentions), name)

head(station_frequency)
```

##10 most popular stations
```{r, fig.align='left', fig.height=5, fig.width=8, echo = FALSE, warning=FALSE}
station_frequency%>%
  top_n(n = 20, wt = total_mentions)%>%
  ggplot(aes(x = reorder(name, total_mentions), y = n_records, fill = direction))+
  geom_col(position = 'dodge')+
  scale_fill_brewer(palette = 'Set1')+
  scale_y_continuous(expand = c(0, 0))+
  labs(title = 'How often is each station mentioned in the data?',
       subtitle = 'Top stations 10 subselection',
       y = '# mentions in data')+
  theme_bw()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour = 'grey40', size = rel(0.4)),
        panel.grid.minor.y = element_line(colour = 'grey80', size = rel(0.4)),
        legend.position = 'top',
        legend.justification = 'left')
```


##How many records do we have of a particular connection from one station to another?
####For stations that are mentioned at least 30 times as origin and 30 times as destination
```{r, fig.align='left', fig.height=7, fig.width=9, echo=FALSE, warning=FALSE}

occupancy_value_mapping = data_frame(occupancy = c('low', 'medium', 'high'),
                                     occupancy_value = c(1, 2, 3))

from_to_data <- train_data%>%
  select(querytime, from, to, occupancy)%>%
  inner_join(station_data%>%
             select(uri, station_from = name),
           by = c('from' = 'uri'))%>%
  inner_join(station_data%>%
             select(uri, station_to = name),
           by = c('to' = 'uri'))%>%
  select(-to, -from)%>%
  inner_join(occupancy_value_mapping, 
             by = 'occupancy')

from_to_data%>%
  count(station_from, station_to)%>%
  group_by(station_from)%>%
  filter(sum(n) > 30)%>%
  group_by(station_to)%>%
  filter(sum(n) > 30)%>%
  ungroup()%>%
  complete(station_from, station_to)%>%
  ggplot(aes(x = station_from, y = station_to, fill = n, label = n))+
  geom_tile(colour = 'grey80')+
  geom_text()+
  labs(x = 'Station from',
       y = 'Station to')+
  scale_fill_viridis('# Trains')+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        legend.key.width = grid::unit(3.6, "cm"),
        legend.position = 'top',
        legend.justification = 'left')+
  guides(fill = guide_colorbar(title.position = "top"))
```
Note that there are multiple records where the **station from** equals the **station to**



##What are the busiest hours during the workweek and the weekend?
```{r, fig.align='left', fig.height=3, fig.width=8, echo=FALSE, warning=FALSE}
from_to_data%>%
  mutate(date = floor_date(querytime, unit = 'day'),
         workday = factor(wday(date) %in% 2:6,
                          levels = c(TRUE, FALSE),
                          labels = c('Workday', 'Weekend')),
         hour = hour(round_date(querytime, unit = 'hour')))%>%
  group_by(workday, hour)%>%
  summarise(mean_occupancy = mean(occupancy_value, na.rm = T),
            n_records = n())%>%
  ungroup()%>%
  complete(workday, hour, fill = list(mean_occupancy = 0, n_records = 0))%>%
  # set rare values to 0
  mutate(mean_occupancy = ifelse(n_records < 10,
                                 0,
                                 mean_occupancy))%>%
  ggplot(aes(x = hour, y = mean_occupancy, fill = workday))+
  geom_col(position = "dodge")+
  labs(x = 'Hour of the day',
       y = 'Mean occupancy (1 = low, 3 = high)')+
  scale_fill_brewer(palette = 'Set1')+
  scale_x_continuous(breaks = 0:23)+
  scale_y_continuous(expand = c(0,0))+
  theme_minimal()+
  theme(legend.title = element_blank(),
        legend.position = 'top',
        legend.justification = 'left',
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
  
```

The morning rush hour clearly lasts from 6 until 9 AM, the evening rush is more spread out.

##What is the mean occupancy from one trainstation to another (color fill) and how often was this connection observed (number)
```{r, fig.align='left', fig.height=7, fig.width=8, echo=FALSE, warning=FALSE}
from_to_data%>%
  mutate(date = floor_date(querytime, unit = 'day'))%>%
  group_by(station_from, station_to)%>%
  summarise(mean_occupancy = mean(occupancy_value, na.rm = T),
            n_records = n())%>%
  group_by(station_from)%>%
  filter(sum(n_records) > 30)%>%
  group_by(station_to)%>%
  filter(sum(n_records) > 30)%>%
  ungroup()%>%
  complete(station_from, station_to)%>%
  ggplot(aes(x = station_from, y = station_to, fill = mean_occupancy, label = n_records))+
  geom_tile(colour = 'grey80')+
  geom_text()+
  labs(x = 'Station from',
       y = 'Station to')+
  scale_fill_viridis('Mean occupancy (1 = low, 3 = high)')+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        legend.key.width = grid::unit(3.6, "cm"),
        legend.position = 'top',
        legend.justification = 'left')+
  guides(fill = guide_colorbar(title.position = "top"))
  
```


##How does this differ between workdays and weekends?
```{r, fig.align='left', fig.height=12, fig.width=8, echo=FALSE, warning=FALSE}
from_to_data%>%
  mutate(date = floor_date(querytime, unit = 'day'),
         workday = factor(wday(date) %in% 2:6,
                          levels = c(TRUE, FALSE),
                          labels = c('Workday', 'Weekend')))%>%
  group_by(workday, station_from, station_to)%>%
  summarise(mean_occupancy = mean(occupancy_value, na.rm = T),
            n_records = n())%>%
  group_by(station_from)%>%
  filter(sum(n_records) > 30)%>%
  group_by(station_to)%>%
  filter(sum(n_records) > 30)%>%
  ungroup()%>%
  complete(workday, station_from, station_to)%>%
  ggplot(aes(x = station_from, y = station_to, fill = mean_occupancy, label = n_records))+
  geom_tile(colour = 'grey80')+
  geom_text()+
  facet_grid(workday~.)+
  labs(x = 'Station from',
       y = 'Station to')+
  scale_fill_viridis('Mean occupancy (1 = low, 3 = high)')+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        legend.key.width = grid::unit(3, "cm"),
        legend.position = 'top',
        legend.justification = 'left')+
  guides(fill = guide_colorbar(title.position = "top"))
  
```


##What about rush hour?
###If we assume that rush hour last from 5.30 to 9.30 in the morning and from 16.30 to 19.30 in the evening (see analysis above)
```{r, fig.align='left', fig.height=12, fig.width=8, echo=FALSE, warning=FALSE}
from_to_data%>%
  mutate(date = floor_date(querytime, unit = 'day'),
         workday = factor(wday(date) %in% 2:6,
                          levels = c(TRUE, FALSE),
                          labels = c('Workday', 'Weekend')),
         rush_hour = factor(hour(round_date(querytime, unit = 'hour')) %in% c(6, 7, 8, 9, 17, 18, 19),
                            levels = c(TRUE, FALSE),
                            labels = c('During rush hours', 'Outside rush hours')))%>%
  group_by(rush_hour, station_from, station_to)%>%
  summarise(mean_occupancy = mean(occupancy_value, na.rm = T),
            n_records = n())%>%
  group_by(station_from)%>%
  filter(sum(n_records) > 30)%>%
  group_by(station_to)%>%
  filter(sum(n_records) > 30)%>%
  ungroup()%>%
  complete(rush_hour, station_from, station_to)%>%
  ggplot(aes(x = station_from, y = station_to, fill = mean_occupancy, label = n_records))+
  geom_tile(colour = 'grey80')+
  geom_text()+
  facet_grid(rush_hour~.)+
  labs(x = 'Station from',
       y = 'Station to')+
  scale_fill_viridis('Mean occupancy (1 = low, 3 = high)')+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        legend.key.width = grid::unit(3, "cm"),
        legend.position = 'top',
        legend.justification = 'left')+
  guides(fill = guide_colorbar(title.position = "top"))
  
```


##What if we combine workdays vs weekends and rush hour?
```{r, fig.align='left', fig.height=12, fig.width=15, echo=FALSE, warning=FALSE}
from_to_data%>%
  mutate(date = floor_date(querytime, unit = 'day'),
         workday = factor(wday(date) %in% 2:6,
                          levels = c(TRUE, FALSE),
                          labels = c('Workday', 'Weekend')),
         rush_hour = factor(hour(round_date(querytime, unit = 'hour')) %in% c(7, 8, 16, 17),
                            levels = c(TRUE, FALSE),
                            labels = c('During rush hours', 'Outside rush hours')))%>%
  group_by(workday, rush_hour, station_from, station_to)%>%
  summarise(mean_occupancy = mean(occupancy_value, na.rm = T),
            n_records = n())%>%
  group_by(station_from)%>%
  filter(sum(n_records) > 30)%>%
  group_by(station_to)%>%
  filter(sum(n_records) > 30)%>%
  ungroup()%>%
  complete(workday, rush_hour, station_from, station_to)%>%
  ggplot(aes(x = station_from, y = station_to, fill = mean_occupancy, label = n_records))+
  geom_tile(colour = 'grey80')+
  geom_text()+
  facet_grid(workday~rush_hour)+
  labs(x = 'Station from',
       y = 'Station to')+
  scale_fill_viridis('Mean occupancy (1 = low, 3 = high)')+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = -90, vjust = 0.2, hjust = 0),
        legend.key.width = grid::unit(3, "cm"),
        legend.position = 'top',
        legend.justification = 'left')+
  guides(fill = guide_colorbar(title.position = "top"))
  
```